---
title: 'STAT 447C: Final Project'
author: "Andrew Tran"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse)
library(LaplacesDemon)
library(reshape2)
library(ggplot2)
```

## Introduction

## Goals of the project

## Data description

```{r, echo=FALSE}
df <- read.csv("data/breast_cancer_wisconsin.csv")
df <- as_tibble(df)
```

Our response variable is `Diagnosis`, which is categorical. Currently, it's being labelled as **M** (malignant) or **B** (benign). We want to transform it into binary values so that **1** and **0** are equivalent to **M** and **B** respectively.

```{r, echo=FALSE}
df$Diagnosis <- factor(df$Diagnosis)
df$Diagnosis <- as.numeric(df$Diagnosis) - 1
```

The Breast Cancer Wisconsin (Diagnostic) data set has 30 explanatory variables. However, in fact, they have 10 main features. For each feature, the mean, standard error, and worst/largest  values are recorded. For example, for a single observation in the data, `radius1` is the mean of distances from center to points on the perimeter, `radius2` is the standard error those distances, and `radius3` is largest distance measured. For the sake of length, we are only interested the means.

```{r, echo=FALSE}
df <- df[, 1:11]
```

These ten real-valued features are measurements taken and computed for each cell nucleus:

\begin{center}
  \begin{tabular}{|c c c|} 
    \hline
    Variable & Type & Description \\
    \hline\hline
    Radius & Quantitative & Distance from center to points on the perimeter of the tumor \\ 
    \hline
    Texture & Quantitative & Gray-scale value \\
    \hline
    Perimeter & Quantitative & Size of the tumor\\
    \hline
    Area & Quantitative & Area of the tumor \\
    \hline
    Smoothness & Quantitative & Local variation in radius length \\
    \hline
    Compactness & Quantitative & $\frac{\text{perimeter}^2}{\text{area}} - 1$ \\
    \hline
    Concavity & Quantitative & Severity of concave portions of the contour\\
    \hline
    Concave points & Quantitative & Number of concave portions of the contour\\
    \hline
    Symmetry & Quantitative & Symmetrical measurement of the tumor\\
    \hline
    Fractal dimension & Quantitative & $\text{coastline approximation} - 1$\\
    \hline
  \end{tabular}
\end{center}

## Exploratory Data Analysis

```{r, echo=FALSE}
par(mar = c(4, 3.8, 1, 1))
par(mfrow = c(2,2))
hist(df$radius1, xlab = "radius", main = "")
hist(df$texture1, xlab = "texture", main = "")
hist(df$perimeter1, xlab = "perimeter", main = "")
hist(df$area1, xlab = "area", main = "")
hist(df$smoothness1, xlab = "smoothness", main = "")
hist(df$compactness1, xlab = "compactness", main = "")
hist(df$concavity1, xlab = "concavity", main = "")
hist(df$concave_points1, xlab = "concave points", main = "")
hist(df$symmetry1, xlab = "symmetry", main = "")
hist(df$fractal_dimension1, xlab = "fractal dimension", main = "")
```

```{r, echo=FALSE}
cor_mat <- round(cor(df[, 2:11]), 2)
melted_cor_mat <- melt(cor_mat)
ggplot(data = melted_cor_mat, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() +
  geom_text(aes(Var2, Var1, label = value), color = "black", size = 4) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("") + ylab("") + ggtitle("Correlation Matrix")
```

## Frequentist logistic regression

```{r}
frequentist.naive_reg <- glm(Diagnosis ~ ., data = df, family = binomial(link = "logit"))
```

## Bayesian 

## Appendix

```{r}
summary(frequentist.naive_reg)
```

